### 2017-10-16 mcmc.tex
troubleshooting

Not being mathematicians, we don't know the precise definition of a
pdf\note{Seriously.  It has something to do with measure theory.}, but for our purposes here a pdf is
any single-valued (scalar) function that is non-negative everywhere
(the entire domain of $\pars$), and obeys a normalization condition
\begin{eqnarray}
0 &\leq& p(\pars) \quad \mbox{for all $\pars$}
\\ \label{eq:normp}
1 &=& \int p(\pars)\,\dd\pars
\quad ,
\end{eqnarray}
where, implicitly, the integral is over the full domain of $\pars$.
Importantly, although $p(\pars)$ is single-valued, $\pars$ can be
a multi-element vector or list or blob; it can be arbitrarily large
and complicated.
In data-analysis contexts, $\pars$ will often be the full blob of free
parameters in the model.
Implicitly, the integral in \equationname~(\ref{eq:normp}) is high-dimensional; it
has as many dimensions as there are elements or entries or components of $\pars$.
Also, if there are elements or entries or components of $\pars$ that are discrete
(that is, take on only integer values or equivalent),
then along those dimensions the integral becomes a discrete sum.
This latter is a detail to which we return below. \textbf{HOGG DO WE?}

@@

But it is worth remembering that---just as \MCMC\ is not a good optimizer
(generically, samples will not lie close to the maximum of the posterior pdf)---%
it is also not a good \emph{search algorithm}.
There is no sense in which (standard) \MCMC\ methods are efficient at
searching, or engineered to search, all of parameter space.
If you really need to check all of parameter space, you should put
\MCMC\ aside, and do an exhaustive search.\note{You might want to open a
  bottle of wine or two while the code runs.}

@@

Another important functional test is to \emph{sample the prior pdf}.
This tests the sampler,
  tests that your priors really are what you think they are
  (and there are so many ways for them \emph{not} to be\note{It is
    easy to think you have flat priors when in practice your sampler
    has flat priors in the logarithms of the parameters or inverses of
    the parameters.  Also, it is easy to have improper priors when you
    think they are proper, because limits are not working, or you were
    wrong about the convergence of some integral.  Sometimes the prior pdf in practice has
    different edges than you think because there is some censoring
    happening that you haven't considered.  And then, of course, it is
    possible for your sampler itself to be buggy.}), and tests that
  your prior pdfs are proper (functions that can be normalized).
This test is most effective as a true end-to-end test of your model
  if you structure your code such that the log probability function \code{lnprob}
  input to the \MCMC\ scheme
  is set up as a sum of a log likelihood function \code{lnlikelihood}
  and a prior function \code{lnprior}.
The sampler can be switched to sampling the prior by a simple replacement
  of the \code{lnlikelihood} function
  with a trivial function that always returns zero.

\MCMC\ should run well in this flat-likelihood case (priors are usually easy to sample)
  and the output sample histograms should look like what you expect
  given the prior pdfs you coded.
If the walkers run off to positive or negative infinity,
  and nothing seems to converge,
  your priors are probably not proper.
If the priors don't look as you expect,
  you either have a bug in your code,
  or else a think-o about how your priors ought to look.
Either way, diagnosis in order.

@@

If things are giving trouble and you suspect the likelihood function,
  another important test is to visualize slices through parameter space.
When you call the likelihood function on a grid in each parameter
  (with the others fixed, say, to reasonable values),
  you should see a likelihood value that is a smooth function of each parameter,
  or at least as smooth as you expect.
Things like numerical integration,
  truncated expansions,
  or adaptive approximations
  inside likelihood functions
  can make the function noisy or jagged at small scales in parameter space.
These problems hurt optimizers more than samplers,
  but they usually point to code issues.
In general, plotting the likelihood function as a function of parameter along
  slices in parameter space often reveals bugs or think-os.

>>>>

Not being mathematicians, we don't know the precise definition of a
pdf\note{Seriously.  It has something to do with measure theory.}, but for our purposes here a pdf is
any single-valued (scalar) function that is non-negative everywhere
(the entire domain of $\pars$), and obeys a normalization condition
\begin{eqnarray}
0 &\leq& p(\pars) \quad \mbox{for all $\pars$}
\\ \label{eq:normp}
1 &=& \int p(\pars)\,\dd\pars
\quad ,
\end{eqnarray}
where, implicitly, the integral is over the full domain of $\pars$.
Importantly, although $p(\pars)$ is single-valued, $\pars$ can be
a multi-element vector or list or blob; it can be arbitrarily large
and complicated.
In data-analysis contexts, $\pars$ will often be the full blob of free
parameters in the model.
Implicitly, the integral in \equationname~(\ref{eq:normp}) is high-dimensional; it
has as many dimensions as there are elements or entries or components of $\pars$.
Also, if there are elements or entries or components of $\pars$ that are discrete
(that is, take on only integer values or equivalent),
then along those dimensions the integral becomes a discrete sum.
This latter is a detail to which we return below (briefly, in
  \sectionname~\ref{sec:trouble}).

@@

But it is worth remembering that---just as \MCMC\ is not a good optimizer
(generically, samples will not lie close to the maximum of the posterior pdf)---%
it is also not a good \emph{search algorithm}.
There is no sense in which (standard) \MCMC\ methods are efficient at
searching, or engineered to search, all of parameter space.
If you really need to check all of parameter space, you should put
\MCMC\ aside, and do an exhaustive search (which will in general take an enormous
  amount of time).

@@

Another important functional test is to \emph{sample the prior pdf}.
This tests the sampler,
  tests that your priors really are what you think they are
  (and there are so many ways for them \emph{not} to be\note{It is
    easy to think you have flat priors when in practice your sampler
    has flat priors in the logarithms of the parameters or inverses of
    the parameters.  Also, it is easy to have improper priors when you
    think they are proper, because limits are not working, or you were
    wrong about the convergence of some integral.  Sometimes the prior pdf in practice has
    different edges than you think because there is some censoring
    happening that you haven't considered.  And then, of course, it is
    possible for your sampler itself to be buggy.}), and tests that
  your prior pdfs are proper (functions that can be normalized).
This test is most effective as a true end-to-end test of your model
  if you structure your code such that the log probability function \code{ln_f}
  input to the \MCMC\ scheme
  is set up as a sum of a log-likelihood function \code{ln_likelihood}
  and a log-prior function \code{ln_prior}.
The sampler can be switched to sampling the prior by a simple replacement
  of the \code{ln_likelihood} function
  with a trivial function that always returns zero.

\MCMC\ should run well in this flat-likelihood case (priors are usually easy to sample)
  and the output sample histograms should look like what you expect
  given the prior pdfs you coded.
If the walkers run off to positive or negative infinity,
  and nothing seems to converge,
  your priors are probably not proper.
If the priors don't look as you expect,
  you either have a bug in your code,
  or else a think-o about how your priors ought to look.
Either way, diagnosis is in order.

@@

If things are giving trouble and you suspect the likelihood function,
  another important test is to visualize slices through parameter space.
When you call the likelihood function on a grid in each parameter
  (with the others fixed, say, to reasonable values),
  you should see a likelihood value that is a smooth function of each parameter,
  or at least as smooth as you expect.
Things like numerical integration,
  truncated expansions,
  or adaptive approximations
  inside likelihood functions
  can make the function noisy or jagged at small scales in parameter space.
These problems hurt optimizers more than samplers,
  but they usually point to code issues.
In general, plotting the likelihood function as a function of parameters along
  slices in parameter space often reveals bugs or think-os.
